---
title: "Getting Started with EvoR: Genomic Foundation Models in R"
author: "Himanshu"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting Started with EvoR}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## Introduction

**EvoR** provides a streamlined interface to the **Evo2** genomic foundation model. Evo2 is designed to understand the "biological grammar" of DNA, RNA, and proteins. 

This vignette walks you through the three primary modalities of the package:
1. **Generation**: Creating biological sequences.
2. **Scoring**: Calculating log-likelihoods for Variant Effect Prediction (VEP).
3. **Embeddings**: Extracting high-dimensional features for regulatory analysis.

## Setup

First, load the package and set your NVIDIA BioNeMo API key.

```{r setup}
library(EvoR)

# You can set your API key as an environment variable
# Sys.setenv(NVIDIA_API_KEY = "your_actual_api_key")
```

## 1. Sequence Generation and Scoring

To analyze a sequence, use `evo2_query()`. By default, it returns both the generated sequence and the confidence scores (log-probabilities) for each nucleotide.

```{r generation, eval = FALSE}
# A sample DNA sequence
my_seq <- "ATGCGTACGTAGCTAGCTAGCTAGCTAGC"

# Query the model
response <- evo2_query(my_seq, num_tokens = 10)

# Extract the scores for Variant Effect Prediction (VEP)
scores <- evo2_get_score(response)

# Visualize the confidence scores for the first few bases
head(scores)
```

## 2. Extracting Embeddings

Embeddings are the model's internal representation of the DNA. These 4096-dimensional vectors are extremely useful for downstream machine learning tasks, such as predicting enhancer activity or chromatin state in Type 2 Diabetes (T2D) research.

```{r embeddings, eval = FALSE}
# Request a forward pass for the sequence
emb_res <- evo2_query_embeddings(my_seq)

# Extract the matrix (nucleotides x 4096 features)
# Note: This requires the RcppCNPy package
embeddings <- evo2_get_embeddings(emb_res)

# Check the dimensions
dim(embeddings)
```

## 3. T2D Research Workflow

A typical research workflow using `EvoR` for Type 2 Diabetes epigenomics would involve:
1. Identifying a GWAS risk variant.
2. Creating a "Wild-type" and "Mutant" sequence.
3. Comparing the **Scores** of both sequences to see if the mutation disrupts the model's expectation.
4. Using the **Embeddings** to classify if the variant falls within a cryptic regulatory element.

---
For more details, visit the [BDB-Genomics GitHub](https://github.com/BDB-Genomics/EvoR).
